---
title: "Apache Arrow: Contributor Sustainability Analysis"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    self-contained: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(arrow)
library(dplyr)
library(lubridate)
library(purrr)
library(ggplot2)
library(DT)
library(gh)

pr_details <- read_parquet("../data/cache/pr_details.parquet")
issue_details <- read_parquet("../data/cache/issue_details.parquet")
```

## Overview

This report examines contributor sustainability for the Apache Arrow project,
drawing on two key metrics from the
[CHAOSS Practitioner Guide](https://chaoss.community/practitioner-guide-contributor-sustainability/):

1. **Active Contributors** — the size of the contributor pool over time.
2. **Contributor Absence Factor (CAF)** — how concentrated the workload is among
   a small number of people.

Together these give a picture of how resilient the project is to individual
contributors stepping away.

## Activity Over Time

```{r}
#| label: activity-over-time

activity_by_month <- bind_rows(
  pr_details |>
    mutate(date = floor_date(as.Date(created_at), "month"), type = "PRs") |>
    count(date, type, name = "n"),
  issue_details |>
    mutate(date = floor_date(as.Date(created_at), "month"), type = "Issues") |>
    count(date, type, name = "n")
) |>
  filter(date >= floor_date(Sys.Date() - months(18), "month"))

ggplot(activity_by_month, aes(x = date, y = n, colour = type)) +
  geom_line(aes(linetype = "Actual"), linewidth = 0.5) +
  geom_smooth(aes(linetype = "Trend"), method = "loess", span = 0.6, se = FALSE, linewidth = 1) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_colour_manual(
    values = c("PRs" = "#4C72B0", "Issues" = "#E15759"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Actual" = "dotted", "Trend" = "solid"),
    name = NULL
  ) +
  labs(
    title = "PRs and Issues Opened per Month",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

## Active Contributors

The plot below shows the number of unique contributors who opened at least one
PR or issue in a rolling 90-day window, sampled monthly over the past 18 months.

```{r}
#| label: active-contributors

all_activity <- bind_rows(
  pr_details |> select(user_login, created_at),
  issue_details |> select(user_login, created_at)
) |>
  filter(!grepl("\\[bot\\]", user_login, ignore.case = TRUE)) |>
  mutate(date = as.Date(created_at))

months <- seq(floor_date(Sys.Date() - months(18), "month"), Sys.Date(), by = "month")

active_contributors <- map_dfr(months, function(m) {
  window_start <- m - days(90)
  active <- all_activity |>
    filter(date >= window_start, date < m) |>
    distinct(user_login) |>
    nrow()
  tibble(month = m, active_contributors = active)
})

ggplot(active_contributors, aes(x = month, y = active_contributors)) +
  geom_line(linewidth = 1, colour = "#4C72B0") +
  geom_point(size = 2, colour = "#4C72B0") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title = "Active Contributors (90-day rolling window)",
    x = NULL,
    y = "Active contributors"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

A contributor is considered "active" if they opened at least one PR or issue,
in the prior 90 days. Bot accounts are excluded.

## Contributor Absence Factor

The Contributor Absence Factor (CAF) is the **minimum number of contributors
whose absence would remove 50% of all merged PRs** in a given quarter. A low
CAF signals that a large share of the work rests on very few shoulders —
increasing risk if any of them step back.

```{r}
#| label: caf-by-quarter

quarter_starts <- seq(
  floor_date(Sys.Date() - months(18), "quarter"),
  floor_date(as.Date("2026-01-01"), "quarter"),
  by = "quarter"
)

quarter_labels <- paste0("Q", quarter(quarter_starts), " ", year(quarter_starts))

caf_quarterly <- map2_dfr(quarter_starts, quarter_labels, function(qs, ql) {
  qe <- qs + months(3)

  q_data <- pr_details |>
    filter(!grepl("\\[bot\\]", user_login, ignore.case = TRUE)) |>
    filter(!is.na(merged_at)) |>
    filter(merged_at >= qs, merged_at < qe) |>
    count(user_login, name = "prs_merged") |>
    arrange(desc(prs_merged)) |>
    mutate(
      quarter = ql,
      rank    = row_number(),
      pct     = prs_merged / sum(prs_merged) * 100,
      cum_pct = cumsum(pct)
    )

  caf_n <- min(q_data$rank[q_data$cum_pct >= 50])
  q_data |> mutate(is_caf = rank <= caf_n, caf_n = caf_n)
})

caf_quarterly |>
  slice_head(n = 10, by = quarter) |>
  mutate(
    quarter    = factor(quarter, levels = quarter_labels),
    user_login = paste(quarter, user_login),
    user_login = reorder(user_login, pct)
  ) |>
  mutate(label = sub("^Q[0-9] [0-9]+ ", "", user_login)) |>
  ggplot(aes(x = user_login, y = pct, fill = is_caf)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.05, size = 2.5) +
  coord_flip() +
  scale_x_discrete(labels = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  scale_fill_manual(
    values = c("TRUE" = "#E15759", "FALSE" = "#4C72B0"),
    labels = c("TRUE" = "CAF contributors (50% of PRs)", "FALSE" = "Other top contributors"),
    name   = NULL
  ) +
  facet_wrap(~ quarter, scales = "free_y", ncol = 4) +
  labs(
    title    = "Contributor Absence Factor by Quarter",
    subtitle = "Red = fewest contributors accounting for 50% of merged PRs",
    x        = NULL,
    y        = "% of merged PRs"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    strip.text      = element_text(face = "bold")
  )
```

## Who Merges PRs?

While the CAF tells us how concentrated PR *authorship* is, it's equally
important to understand who actually *merges* contributions. Merge authority
sits with a small group of committers — and if that group is too small or too
concentrated, it becomes a bottleneck even when there are plenty of willing
contributors.

```{r}
#| label: fetch-merged-by

merged_by_data <- pr_details |>
  filter(
    !is.na(merged_by),
    !grepl("\\[bot\\]", merged_by, ignore.case = TRUE),
    !is.na(merged_at)
  ) |>
  mutate(merged_at = as.Date(merged_at)) |>
  select(number, merged_by, merged_at)
```

```{r}
#| label: who-merges-plot

merger_quarters <- map2_dfr(quarter_starts, quarter_labels, function(qs, ql) {
  qe <- qs + months(3)

  q_data <- merged_by_data |>
    filter(merged_at >= qs, merged_at < qe) |>
    count(merged_by, name = "prs_merged") |>
    arrange(desc(prs_merged)) |>
    mutate(
      quarter = ql,
      rank    = row_number(),
      pct     = prs_merged / sum(prs_merged) * 100,
      cum_pct = cumsum(pct)
    )

  caf_n <- min(q_data$rank[q_data$cum_pct >= 50])
  q_data |> mutate(is_caf = rank <= caf_n, caf_n = caf_n)
})

merger_quarters |>
  slice_head(n = 10, by = quarter) |>
  mutate(
    quarter   = factor(quarter, levels = quarter_labels),
    merged_by = paste(quarter, merged_by),
    merged_by = reorder(merged_by, pct)
  ) |>
  mutate(label = sub("^Q[0-9] [0-9]+ ", "", merged_by)) |>
  ggplot(aes(x = merged_by, y = pct, fill = is_caf)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.05, size = 2.5) +
  coord_flip() +
  scale_x_discrete(labels = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  scale_fill_manual(
    values = c("TRUE" = "#E15759", "FALSE" = "#4C72B0"),
    labels = c("TRUE" = "CAF mergers (50% of merges)", "FALSE" = "Other top mergers"),
    name   = NULL
  ) +
  facet_wrap(~ quarter, scales = "free_y", ncol = 4) +
  labs(
    title    = "Who Merges PRs? (by quarter)",
    subtitle = "Red = fewest people accounting for 50% of merges that quarter",
    x        = NULL,
    y        = "% of merges"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    strip.text      = element_text(face = "bold")
  )
```

Merge activity is highly concentrated: in most quarters, just a handful of
committers account for half of all merges. Contributors can open PRs freely,
but the ability to land those PRs depends on a small group of committers.

### What the data shows

The CAF dropped sharply from **11 in Q2 2024 to 5 in Q4 2024**, and has held
steady at 5 ever since. In practical terms, this means that just 5 people now
account for half of all merged PRs in any given quarter — down from 11 a year
and a half ago.

A single contributor - **kou** - has been the top merger in every quarter
observed, contributing between roughly 10-25% of all merged PRs. Their share
peaked at 25% in Q3 2025, meaning one person was responsible for a quarter of
all merged work that period.

At the same time, the **total number of contributors merging PRs** has declined
from 94 (Q2 2024) to 59 (Q1 2026). Fewer people are doing the work, and the
work that remains is more concentrated among a smaller core.

### Why this matters

The decline in both the CAF and the total contributor pool has a clear
structural cause: the dissolution of Voltron Data, which employed a significant
number of Arrow committers. When a company that pays people to work full-time
on a project shuts down, those contributors don't disappear overnight, but
their available hours shrink considerably. The data reflects exactly that —
fewer total contributors (94 in Q2 2024 → 59 in Q1 2026), and the work that
remains increasingly concentrated among a smaller group of individuals
employed by other organisations, but typically with narrower Arrow-specific
mandates.

A CAF of 5 sustained over several quarters is manageable, but fragile. The
project is now largely dependent on a small number of people.  
Any further attrition in that group would have
an outsized impact on throughput.

## New Contributors

A healthy project needs a steady pipeline of new contributors — both to grow
the contributor pool and to offset natural attrition. We identify first-time
contributors by finding each person's earliest PR in the repository.

```{r}
#| label: new-contributors-overall

# Each person's first PR
first_prs <- pr_details |>
  filter(!grepl("\\[bot\\]", user_login, ignore.case = TRUE)) |>
  group_by(user_login) |>
  slice_min(created_at, n = 1) |>
  ungroup()

# First PRs within the 12-month window
new_contribs <- first_prs |>
  filter(as.Date(created_at) >= floor_date(Sys.Date() - months(12), "month")) |>
  mutate(month = floor_date(as.Date(created_at), "month"))

# Monthly counts by outcome
monthly_new <- new_contribs |>
  filter(month > as.Date("2025-03-01")) |>
  mutate(
    outcome = case_when(
      !is.na(merged_at) ~ "Merged",
      state == "closed" ~ "Closed",
      TRUE ~ "Open"
    )
  ) |>
  count(month, outcome, name = "n") |>
  mutate(outcome = factor(outcome, levels = c("Open", "Closed", "Merged")))

ggplot(monthly_new, aes(x = month, y = n, fill = outcome)) +
  geom_area(alpha = 0.8) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y") +
  scale_fill_manual(
    values = c("Merged" = "#59A14F", "Closed" = "#E15759", "Open" = "#4C72B0"),
    name = NULL
  ) +
  labs(
    title = "First-Time Contributors per Month",
    subtitle = "Outcome of each person's first PR",
    x = NULL,
    y = "Contributors"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r}
#| label: new-contributors-stats
#| include: false

n_first_timers <- nrow(new_contribs)
n_merged <- sum(!is.na(new_contribs$merged_at))
n_closed <- sum(new_contribs$state == "closed" & is.na(new_contribs$merged_at))
n_open <- sum(new_contribs$state == "open")
pct_merged <- round(n_merged / n_first_timers * 100)
```

Over the past 12 months, **`r n_first_timers` people** opened their first PR to
Apache Arrow. Of those first PRs, **`r n_merged` (`r pct_merged`%)** were merged,
`r n_closed` were closed without merging, and `r n_open` are still open.

### PR Designation

```{r}
#| label: review-setup
#| include: false

review_cache   <- "first_timer_review_status.csv"
bot_pattern    <- "\\[bot\\]|github-actions|codecov|dependabot"
abandoned_days <- 90

parse_activity <- function(issue_comments, pr_reviews) {
  ic <- if (length(issue_comments) > 0) {
    map_dfr(issue_comments, function(c) tibble(
      login = c$user$login,
      at    = as.POSIXct(c$created_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
    ))
  } else tibble(login = character(), at = as.POSIXct(character()))

  rv <- if (length(pr_reviews) > 0) {
    map_dfr(pr_reviews, function(r) tibble(
      login = r$user$login,
      at    = as.POSIXct(r$submitted_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
    ))
  } else tibble(login = character(), at = as.POSIXct(character()))

  bind_rows(ic, rv) |>
    filter(!grepl(bot_pattern, login, ignore.case = TRUE)) |>
    arrange(at)
}
```

```{r}
#| label: review-stage1
#| include: false
# Stage 1: Build CSV from parquet if it doesn't exist or has new PRs.
# new_contribs contains each person's first PR to the repo.

if (!file.exists(review_cache)) {
  review_status <- new_contribs |>
    select(number, author = user_login) |>
    mutate(
      designation = NA_character_,
      first_response_at = as.POSIXct(NA),
      days_to_first_response = NA_real_,
      first_responder = NA_character_
    )
  write.csv(review_status, review_cache, row.names = FALSE)
} else {
  review_status <- read.csv(
    review_cache,
    colClasses = c(number = "integer", author = "character",
                   designation = "character")
  ) |>
    mutate(
      first_response_at      = as.POSIXct(first_response_at, tz = "UTC"),
      days_to_first_response = as.numeric(days_to_first_response)
    )
  missing <- setdiff(new_contribs$number, review_status$number)
  if (length(missing) > 0) {
    new_rows <- new_contribs |>
      filter(number %in% missing) |>
      select(number, author = user_login) |>
      mutate(designation = NA_character_)
    review_status <- bind_rows(review_status, new_rows)
    write.csv(review_status, review_cache, row.names = FALSE)
  }
}
```

```{r}
#| label: review-stage1b
#| include: false
# Stage 1b: Mark PRs less than 6 weeks old as "recent" — too new to classify

recent_numbers <- new_contribs |>
  filter(as.Date(created_at) >= Sys.Date() - 42) |>
  pull(number)

newly_recent <- recent_numbers[
  is.na(review_status$designation[review_status$number %in% recent_numbers])
]

if (length(newly_recent) > 0) {
  message("Stage 1b: marking ", length(newly_recent), " PRs as recent (< 6 weeks old)")
  review_status$designation[review_status$number %in% newly_recent] <- "recent"
  write.csv(review_status, review_cache, row.names = FALSE)
}
```

```{r}
#| label: review-stage2
#| include: false
# Stage 2: Check reviews for PRs with no designation yet — classify as
#          "no-review" if no human has commented, otherwise leave NA for Stage 3

needs_review_check <- review_status |>
  filter(is.na(designation)) |>
  pull(number)

if (length(needs_review_check) > 0) {
  message("Stage 2: checking reviews for ", length(needs_review_check), " PRs...")

  for (pr_number in needs_review_check) {
    author <- review_status$author[review_status$number == pr_number]

    issue_comments <- tryCatch(
      gh("GET /repos/{owner}/{repo}/issues/{issue_number}/comments",
         owner = "apache", repo = "arrow", issue_number = pr_number,
         per_page = 100, .token = Sys.getenv("GITHUB_PAT")),
      error = function(e) list()
    )
    pr_reviews <- tryCatch(
      gh("GET /repos/{owner}/{repo}/pulls/{pull_number}/reviews",
         owner = "apache", repo = "arrow", pull_number = pr_number,
         per_page = 100, .token = Sys.getenv("GITHUB_PAT")),
      error = function(e) list()
    )

    activity       <- parse_activity(issue_comments, pr_reviews)
    human_activity <- activity |> filter(login != author)

    if (nrow(human_activity) == 0) {
      review_status$designation[review_status$number == pr_number] <- "no-review"
    } else {
      # Record first response
      first_response <- human_activity |> slice_head(n = 1)
      pr_created <- new_contribs$created_at[new_contribs$number == pr_number]
      review_status$first_response_at[review_status$number == pr_number] <- first_response$at
      review_status$days_to_first_response[review_status$number == pr_number] <-
        as.numeric(difftime(first_response$at, pr_created, units = "days"))
      review_status$first_responder[review_status$number == pr_number] <- first_response$login
    }

    Sys.sleep(0.2)
  }

  write.csv(review_status, review_cache, row.names = FALSE)
  message("Stage 2 complete.")
}
```

```{r}
#| label: review-stage3
#| include: false
# Stage 3: For reviewed PRs (designation still NA), check if last actor was
#          not the author and last activity > abandoned_days ago → "abandoned",
#          activity in past 4 weeks → "active", otherwise leave NA

needs_abandoned_check <- review_status |>
  filter(is.na(designation)) |>
  pull(number)

if (length(needs_abandoned_check) > 0) {
  message("Stage 3: checking abandonment for ", length(needs_abandoned_check), " PRs...")

  for (pr_number in needs_abandoned_check) {
    author <- review_status$author[review_status$number == pr_number]

    issue_comments <- tryCatch(
      gh("GET /repos/{owner}/{repo}/issues/{issue_number}/comments",
         owner = "apache", repo = "arrow", issue_number = pr_number,
         per_page = 100, .token = Sys.getenv("GITHUB_PAT")),
      error = function(e) list()
    )
    pr_reviews <- tryCatch(
      gh("GET /repos/{owner}/{repo}/pulls/{pull_number}/reviews",
         owner = "apache", repo = "arrow", pull_number = pr_number,
         per_page = 100, .token = Sys.getenv("GITHUB_PAT")),
      error = function(e) list()
    )

    all_events <- parse_activity(issue_comments, pr_reviews)
    last_event <- if (nrow(all_events) > 0) slice_tail(all_events, n = 1) else NULL

    days_since <- if (!is.null(last_event)) {
      as.numeric(Sys.time() - last_event$at, units = "days")
    } else NA_real_

    designation <- if (
      !is.null(last_event) && days_since <= 28
    ) "active" else if (
      !is.null(last_event) &&
      last_event$login != author &&
      days_since > abandoned_days
    ) "abandoned" else "stale-risk"

    review_status$designation[review_status$number == pr_number] <- designation
    Sys.sleep(0.2)
  }

  write.csv(review_status, review_cache, row.names = FALSE)
  message("Stage 3 complete.")
}

# Summary stats
new_contribs_status <- new_contribs |>
  mutate(
    status = case_when(
      !is.na(merged_at) ~ "Merged",
      state == "closed" ~ "Closed (not merged)",
      TRUE              ~ "Open"
    )
  ) |>
  left_join(review_status |> select(number, designation, days_to_first_response),
            by = "number")

n_opened       <- nrow(new_contribs_status)
n_merged       <- sum(new_contribs_status$status == "Merged")
n_closed       <- sum(new_contribs_status$status == "Closed (not merged)")
n_open         <- sum(new_contribs_status$status == "Open")
open_contribs  <- new_contribs_status |> filter(status == "Open")
n_no_review    <- sum(open_contribs$designation == "no-review",  na.rm = TRUE)
n_abandoned    <- sum(open_contribs$designation == "abandoned",  na.rm = TRUE)
n_active       <- sum(open_contribs$designation == "active",     na.rm = TRUE)
n_recent       <- sum(open_contribs$designation == "recent",     na.rm = TRUE)
n_stale_risk   <- sum(open_contribs$designation == "stale-risk", na.rm = TRUE)
pct_no_review  <- round(n_no_review  / n_open * 100)
pct_abandoned  <- round(n_abandoned  / n_open * 100)
pct_stale_risk <- round(n_stale_risk / n_open * 100)

ttfr        <- review_status |> filter(!is.na(days_to_first_response)) |> pull(days_to_first_response)
ttfr_median <- round(median(ttfr), 1)
ttfr_p75    <- round(quantile(ttfr, 0.75), 1)
ttfr_max    <- round(max(ttfr), 0)
```

Of those `r n_open` open PRs:

- **`r n_no_review` (`r pct_no_review`%)** received no human review at all
- **`r n_abandoned` (`r pct_abandoned`%)** are abandoned - reviewed but no
  response from the author in over 90 days
- **`r n_stale_risk` (`r pct_stale_risk`%)** are at risk of becoming stale - have had some
  engagement but no maintainer response in over 6 weeks
- **`r n_active`** are actively being discussed (recent activity in the past 4 weeks)
- **`r n_recent`** were opened in the past 6 weeks

### Time to First Response

For the `r length(ttfr)` PRs that received a human response, the median time
to first response was **`r ttfr_median` days**, with 75% receiving a response
within `r ttfr_p75` days. The slowest response took `r ttfr_max` days.

```{r}
#| label: time-to-first-response

review_status |>
  filter(!is.na(days_to_first_response), days_to_first_response <= 100) |>
  mutate(days_to_first_response = as.numeric(days_to_first_response)) |>
  ggplot(aes(x = days_to_first_response)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", colour = "white") +
  geom_vline(xintercept = ttfr_median, linetype = "dashed",
             colour = "#E15759", linewidth = 0.8) +
  annotate("text", x = ttfr_median + 1, y = Inf,
           label = paste0("median: ", ttfr_median, " days"),
           hjust = 0, vjust = 1.5, colour = "#E15759", size = 3.5) +
  labs(
    title = "Time to First Response (reviewed PRs only)",
    subtitle = "Limited to PRs with response within 100 days",
    x     = "Days to first human response",
    y     = "Number of PRs"
  ) +
  theme_minimal()
```

## Conclusion

**The contributor base has contracted.** Following the dissolution of Voltron Data,
the Contributor Absence Factor fell from 11 to 5: just five people now account for
half of all merged PRs. Merge authority is similarly concentrated.

**The maintainers are doing their part.** First-time contributors get a response
in a median of 1.1 days. Half get their first PR merged. Of the 75 first-timer PRs
still open, only 10 received no review — the rest were reviewed but stalled (17
abandoned by the author, 20 at risk of stalling). The bottleneck isn't attention;
it's that a shrinking group is stretched thin.

**The risk is fragility.** If one or two of the top five contributors step back,
review capacity and release velocity will be directly affected. The project would
benefit from additional maintainer time, whether through sponsorship or dedicated
engineering support from organizations that depend on Arrow.
