--- 
title: "Arrow Maintainer Dashboard"
author: "Alenka Frim and Nic Crane"
format: 
  dashboard:
    logo: ./images/arrow-logo.png
    nav-buttons: 
      - github
      - href: https://github.com/thisisnic/arrowdash/
---

<!-- 

Needs Quarto 1.4 installed to run
Seek inspiration from https://quarto.org/docs/dashboards/examples/

-->

```{r}
#| label: r_setup
#| include: false
library(reticulate)
library(DT)
reticulate::use_virtualenv('r-arrow-dash')
update_time <- format(Sys.time(), "%a %b %d %X %Y %z")
```

Last updated: `{r} update_time`
```{python}
#| label: main methods
#| include: false
#| warning: false

import github_data.data as gd
data = gd.get_data()

import github_data.methods as gha

p_issues, p_prs = gha.get_all(data, "Python")
r_issues, r_prs = gha.get_all(data, "R")

p_issues_open = gha.get_open(p_issues)
r_issues_open = gha.get_open(r_issues)

p_prs_open = gha.get_open(p_prs)
r_prs_open = gha.get_open(r_prs)

p_issues_new_open = gha.get_new_contributors_open(p_issues)
r_issues_new_open = gha.get_new_contributors_open(r_issues)

p_prs_new_open = gha.get_new_contributors_open(p_prs)
r_prs_new_open = gha.get_new_contributors_open(r_prs)

p_issues_sum = gha.get_summary(p_issues)
p_prs_sum = gha.get_summary(p_prs)
r_issues_sum = gha.get_summary(r_issues)
r_prs_sum = gha.get_summary(r_prs)
```

# PyArrow

Issues, issues not yet closed (open issues) and pull requests created in the last 3 months. TODO: Questions?

## Row

```{python}
#| content: valuebox
#| title: "All issues"
dict(
  icon = "envelope",
  color = "light",
  value = len(p_issues)
)
```

```{python}
#| content: valuebox
#| title: "Open Issues"
dict(
  icon = "envelope-open",
  color = "light",
  value = len(p_issues_open)
)
```

```{python}
#| content: valuebox
#| title: "All Pull Requests"
dict(
  icon = "code-square",
  color = "light",
  value = len(p_prs_open)
)
```

```{python}
#| content: valuebox
#| title: "Stack Overflow Questions"
dict(
  icon = "question-circle",
  color = "light",
  value = "2"
)
```

## Row

### Column {width=50%}

```{r}
#| title: List of issues not yet closed
DT::datatable(escape=FALSE,py$p_issues_open)
```

```{r}
#| title: List of issues not yet closed
#| label: py_issues_open
DT::datatable(escape = FALSE, py$p_issues_open)
```


### Column {width=50%}

```{r}
#| label: p_prs_open
#| title: List of Pull Requests not yet merged
DT::datatable(escape=FALSE,py$p_prs_open)
```

## Row

### Column {width=50%}

```{r}
#| title: List of open issues by new contributors
#| label: p_issues_new
DT::datatable(escape=FALSE,py$p_issues_new_open)
```

### Column {width=50%}

```{r}
#| title: List of open Pull Requests by new contributors
#| label: py_prs_new
DT::datatable(escape=FALSE,py$p_prs_new_open)
```

## Row

### Column {width=50%}

```{python}
#| title: Open issues by week
import altair as alt

alt.Chart(p_issues_sum).mark_bar(cornerRadius=10).encode(
    x=alt.X("created_at:N").axis(labelAngle=-45).title("Week"),
    y=alt.Y("labels:Q").title("Number of issues")
)
```

### Column {width=50%}

```{python}
#| title: Open prs by week
alt.Chart(p_prs_sum).mark_bar(cornerRadius=10).encode(
    x=alt.X("created_at:N").axis(labelAngle=-45).title("Week"),
    y=alt.Y("labels:Q").title("Number of prs")
)
```

# Arrow R

Issues, issues not yet closed (open issues) and pull requests created in the last 3 months. TODO: Questions?

## Row

```{python}
#| content: valuebox
#| title: "All Issues"
dict(
  icon = "envelope",
  color = "light",
  value = len(r_issues)
)
```

```{python}
#| content: valuebox
#| title: "Open Issues"
dict(
  icon = "envelope-open",
  color = "light",
  value = len(r_issues_open)
)
```

```{python}
#| content: valuebox
#| title: "All Pull Requests"
dict(
  icon = "code-square",
  color = "light",
  value = len(r_prs_open)
)
```

```{python}
#| content: valuebox
#| title: "Stack Overflow Questions"
dict(
  icon = "question-circle",
  color = "light",
  value = "2"
)
```

## Row

### Column {width=50%}

```{r}
#| title: List of issues not yet closed
DT::datatable(escape=FALSE,py$r_issues_open)
```

### Column {width=50%}

```{r}
#| title: List of Pull Requests not yet merged
DT::datatable(escape=FALSE,py$r_prs_open)
```

## Row

### Column {width=50%}

```{r}
#| title: List of open issues by new contributors
DT::datatable(escape=FALSE,py$r_issues_new_open)
```

### Column {width=50%}

```{r}
#| title: List of open Pull Requests by new contributors
DT::datatable(escape=FALSE,py$r_prs_new_open)
```

## Row

### Column {width=50%}

```{python}
#| title: Open issues by week
alt.Chart(r_issues_sum).mark_bar(cornerRadius=10).encode(
    x=alt.X("created_at:N").axis(labelAngle=-45).title("Week"),
    y=alt.Y("labels:Q").title("Number of issues")
)
```

### Column {width=50%}

```{python}
#| title: Open prs by week
alt.Chart(r_prs_sum).mark_bar(cornerRadius=10).encode(
    x=alt.X("created_at:N").axis(labelAngle=-45).title("Week"),
    y=alt.Y("labels:Q").title("Number of prs")
)
```
